\thispagestyle{plain}

\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration}

% \vspace{20pt}

% Masters (Research)
By submitting this project electronically, I declare that the entirety of the work contained
therein is my own, original work, that I am the sole author thereof (save to the extent explicitly
otherwise stated), that reproduction and publication thereof by Stellenbosch University will
not infringe any third party rights and that I have not previously in its entirety or in part
submitted it for obtaining any qualification.

In the preparation of this project, artificial intelligence (AI) tools were utilised solely as supportive instruments to enhance the quality and efficiency of the research process. The use of AI was limited to improving the clarity, structure, and grammatical quality of the written content. All technical content, experimental design, analysis, and conclusions presented are entirely the result of my own work, understanding, and interpretation. Any text generated through AI assistance were critically reviewed, revised, and restructured to ensure full alignment with my personal voice, intended meaning, and academic integrity.

With regard to coding, AI tools were employed to expedite routine programming tasks such as generating boilerplate code, refactoring existing code for improved readability, and identifying potential bugs. All AI-assisted code was carefully validated, modified where necessary, and verified to behave as intended. The use of AI in this context served only to improve workflow efficiency and did not influence the originality or integrity of the implemented methods or results.

\vspace{35pt}
% \vspace{1cm}
\noindent
\begin{minipage}{.5\textwidth}
    \noindent
    \phantom{Date:}~\hfill\makebox[0pt][c]{November 2025}\hfill\mbox{}\\[-.5\baselineskip]
    Date:~ \dotfill\mbox{}\par
\end{minipage}

\vspace{35pt}

\vfill

% \begin{center}
    Copyright © 2025 Stellenbosch University\par
    All rights reserved
% \end{center}

\vspace{35pt}

